---
title: "2015-2016 NFL Football Season - Making A Beautiful Scatter Plot with R"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

This is a step-by-step breakdown of my thought process for building a nice looking scatter plot using `ggplot2` in R. This analysis uses a dataset containing information on every play in the 2015-2016 NFL football season found [here on Kaggle](https://www.kaggle.com/maxhorowitz/nflplaybyplay2015). This season culminated in Super Bowl L in which the Denver Broncos defeated the Carolina Panthers 24-10. It was also the last professional football game for the legendary [Peyton Manning](https://en.wikipedia.org/wiki/Peyton_Manning). Not a bad way to go out.

In this exploration I'll take a look at some simple summary statistics of the various teams over the regular season and see if any interesting insights come out as to what made the Broncos/Panthers the final two standing for the year. I'll build a scatter plot comparing the total offense and defensive production of each team and use some simple web scraping to add the team colors to the visualization.

## Data Analysis 
First, load the relevant packages. I use the usual `dplyr` and `tidyr` for data frame manipulation, `magrittr` for the two-way chaining operator,  `lubridate` for working with dates, and `ggplot2` for visualizations.

```{r echo=TRUE, message=FALSE, warning=FALSE}

library(tidyr)
library(dplyr)
library(magrittr)
library(lubridate)
library(ggplot2)
```

Read in the CSV file and let's see what we have

```{r}
df <- data.frame(read.csv('/Users/ajpryor/Downloads/NFLPlaybyPlay2015.csv',header=T, stringsAsFactors = FALSE))
df$Date <- as.Date(df$Date)
str(df)
```

There's a ton of information here. It seems every play of every game in the regular season is catalogued. Each game has a `GameID` and the `Date` on which it was played. The offense and defense teams are given by `posteam` and `DefensiveTeam`, respectively. There's also the time on the clock, current down and distance, type of play, special teams information, information on the players involved, etc. It's really tempting to explore this in great detail, but for the purposes of this demonstration, let's keep it simple. I'll organize things a bit and create some summary statistics. 

NFL games usually, but not always, are played on Sundays, so they are instead grouped by week. First, I'll assign the week number to each game relative to the first game of the season.

```{r}

df %<>%
  mutate(week = as.numeric(df$Date - df[1,'Date'])%/%7 + 1)
```


Let's breakdown the overall offensive/defensive performance for each time to see which was the driving factor behind the Bronco's victory.
To start I'll compare the total yards of offense per game, the total yards given up defensively per game, and the average number of points scored per game. We can get to these values by grouping the data by `posteam` or `DefensiveTeam` and aggregating `Yards.Gained` to get the total number of yards in the season. Dividing by the number of unique `GameID`'s for each team gets the average (just in case some teams played more games than others). 

Defense first as it is a little simpler:

```{r}
def <- df %>% 
  filter(!is.na(DefensiveTeam)& DefensiveTeam!='') %>%
  group_by(DefensiveTeam) %>% 
  summarise(def_ypg = sum(Yards.Gained[Penalty.Yards==0]) / n_distinct(GameID)) %>%
  arrange(def_ypg)
def <- data.frame(def)
def
```

Spoiler alert -- Denver had an amazing defense this year.

Now for the offense.
The total yards per game is calculated essentially the same as for defense, but the total number of points scored per game (ppg) requires a little manipulation, as teams can score in any of the following ways:

  - Touchdown (6 points)
  - Extra Point (1 point)
  - Two-point Conversion (2 points)
  - Field-Goal (3 points)

In R this can be done as

```{r}
off <- df %>% 
  filter(!is.na(posteam) & posteam!='') %>%
  group_by(posteam) %>% 
  summarise(off_ypg = sum(Yards.Gained[Penalty.Yards==0]) / n_distinct(GameID),
            ppg=(   sum(Touchdown*6, na.rm=TRUE) +
                    sum(ifelse(ExPointResult=="Success",1,0),na.rm=TRUE) + 
                    sum(ifelse(TwoPointConv=="Success",2,0),na.rm=TRUE)  +  
                    sum(ifelse(FieldGoalResult=="Good",3,0),na.rm=TRUE))/ n_distinct(GameID)) %>%
  arrange(desc(ppg))                                                 
off <- data.frame(off)
off
```

If you're fact checking me you may wonder why these numbers are different than what was reported for the year (i.e. the [Cardinals were #2 in ppg, not #3](http://www.pro-football-reference.com/teams/crd/2015.htm)). However, note I've only calculated the points *scored by the offense*. The missing points are from kickoff returns for TD, pick-sixes, safeties, etc.


Now let's combine these two to get the stats for each time in one place

```{r}
nfl <- def %>%
  inner_join(off,by=c('DefensiveTeam'='posteam')) %>%
  rename(Team=DefensiveTeam) %>%
  arrange(Team)
nfl

```

Let's make a scatter plot with defense on one axis and offense on the other. To add an extra dimension, I'll make the size of the points indicate the average points per game

```{r}
plt <- ggplot(data=nfl,aes(x=off_ypg,y=def_ypg)) + geom_point(aes(color=Team, size=ppg)) +
   labs(title="2015-2016 NFL Offense/Defense Overview",x="Offensive YPG", y="Defensive YPG")
plt

```

That's an okay start, but the default colors, although pretty, are kind of confusing in this context. It would look a lot better if we colored each data point using the actual team colors. Time for some simple web scraping. 

## Coloring Data by Team Color

Google led me to find a list of the hex codes for each team's colors [here](http://dynasties.operationsports.com/team-colors.php?sport=nfl). Using `htmltab`, let's try to extract this information from the website.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(htmltab)
url <- "http://dynasties.operationsports.com/team-colors.php?sport=nfl"
nfl.colors <- htmltab(url)
nfl.colors
```

`htmltab` has a lot more sophisticated features for parsing website data, but after seeing the output in this form I knew I could work with it so I didn't bother digging any deeper into their documentation. I'll just ignore the rows for the All-Stars teams, and since I want to use the team's primary color I'll discard any black, white, or gray colors with a little regular expression magic.

```{r}
nfl.colors <- nfl.colors[1:32,]
nfl.colors <- gsub("#FFFFFF|#000000|(#D[0-9]D[0-9]D[0-9])","",nfl.colors)
nfl.colors <- data.frame(data=nfl.colors)
head(nfl.colors)
```

At this point some teams have only one color, some have several, and for a few of them the color I would choose is backwards. First let's split this string into columns, and fill in the blanks with duplicated colors.

```{r echo=TRUE, message=FALSE, warning=FALSE}
nfl.colors %<>% separate(col=data,into=c("Team","ThirdColor","SecondColor","FirstColor"),sep="#")
nfl.colors$FirstColor[is.na(nfl.colors$FirstColor)] <- nfl.colors$SecondColor[is.na(nfl.colors$FirstColor)]
nfl.colors$FirstColor[is.na(nfl.colors$FirstColor)] <- nfl.colors$ThirdColor[is.na(nfl.colors$FirstColor)]
nfl.colors$SecondColor[is.na(nfl.colors$SecondColor)] <- nfl.colors$FirstColor[is.na(nfl.colors$SecondColor)]
nfl.colors %<>% mutate(FirstColor=paste("#",FirstColor,sep="")) %>%
  mutate(SecondColor=paste("#",SecondColor,sep="")) %>%
  mutate(ThirdColor=paste("#",ThirdColor,sep="")) %>%
  arrange(Team)
head(nfl.colors)
```

Now I'll match the color with the stats I calculated earlier by giving each team a temporary id by alphabetical order and joining.

```{r}
nfl.colors %<>% mutate(id=1:nrow(nfl.colors))
nfl %<>% mutate(id=1:nrow(nfl)) %>%
  left_join(y=nfl.colors, by="id") %>%
  select(Team=Team.x, def_ypg, off_ypg, ppg,FirstColor,SecondColor,ThirdColor)
head(nfl)
```

Great! Well, almost. Putting the abbreviations and full names both in alphabetical order means that 'Seattle' and 'San Francisco' are in a different order than 'SF' and 'SEA'. Easy to fix.

```{r}
tmp.color.SF <- nfl$FirstColor[nfl$Team=="SF"]
tmp.color.SEA <- nfl$FirstColor[nfl$Team=="SEA"]
nfl$FirstColor[nfl$Team=="SF"]  <- tmp.color.SEA
nfl$FirstColor[nfl$Team=="SEA"] <-tmp.color.SF

tmp.color.SF <- nfl$SecondColor[nfl$Team=="SF"]
tmp.color.SEA <- nfl$SecondColor[nfl$Team=="SEA"]
nfl$SecondColor[nfl$Team=="SF"]  <- tmp.color.SEA
nfl$SecondColor[nfl$Team=="SEA"] <-tmp.color.SF

tmp.color.SF <- nfl$ThirdColor[nfl$Team=="SF"]
tmp.color.SEA <- nfl$ThirdColor[nfl$Team=="SEA"]
nfl$ThirdColor[nfl$Team=="SF"]  <- tmp.color.SEA
nfl$ThirdColor[nfl$Team=="SEA"] <-tmp.color.SF
```

Almost there. For a few teams, the color that has ended up as `FirstColor` is not the one I want to use, so I'll manually switch them.

```{r}
color.scheme <- nfl$FirstColor
swap.color.T <- c("DEN","HOU","CHI","SEA")
swap.color.S <- c("TEN")
color.scheme[nfl$Team %in% swap.color.S] <- nfl$SecondColor[nfl$Team %in% swap.color.S]
color.scheme[nfl$Team %in% swap.color.T] <- nfl$ThirdColor[nfl$Team %in% swap.color.T]
```

Finally, let's apply this color scheme to our plot. The aesthetic color palette in `ggplot2` can be controlled using `scale_color_manual`.
```{r}
plt <- ggplot(data=nfl,aes(x=off_ypg,y=def_ypg)) + geom_point(aes(color=Team, size=ppg)) +
  labs(title="2015-2016 NFL Offense/Defense Overview",x="Offensive YPG", y="Defensive YPG")  + scale_color_manual(values=color.scheme) 
plt
```

Awesome, it worked. Now to touch up this plot to make it look nice.

## Plot Customization

With so many different teams and colors, let's remove the color legend in favor of text labels. The labels can be added with a `geom_text`

```{r}
plt <- ggplot(data=nfl,aes(x=off_ypg,y=def_ypg)) + geom_point(aes(color=Team, size=ppg)) +
  labs(title="2015-2016 NFL Offense/Defense Overview",x="Offensive YPG", y="Defensive YPG")  + scale_color_manual(values=color.scheme) +
  geom_text(aes(label=Team),size=3.0) 
plt
```


The bubbles are too small for some datapoints (poor 49ers....). I'll make things a bit bigger.

```{r echo=TRUE, message=FALSE, warning=FALSE, warning=FALSE, fig.width=10, fig.height=10}
plt <- ggplot(data=nfl,aes(x=off_ypg,y=def_ypg)) + geom_point(aes(color=Team, size=ppg)) +
  labs(title="2015-2016 NFL Offense/Defense Overview",x="Offensive YPG", y="Defensive YPG")  + scale_color_manual(values=color.scheme) +
  geom_text(aes(label=Team),size=3.0)  + 
  xlim(350, 460) + ylim(310,470) +  scale_size(range=c(5,17))
plt
```

I'd prefer the background to be clean, so let's remove the grid lines and use `theme_bw()` to make the background white.

```{r fig.width=10, fig.height=10, echo=TRUE,message=FALSE, warning=FALSE}
plt <- ggplot(data=nfl,aes(x=off_ypg,y=def_ypg)) + geom_point(aes(color=Team, size=ppg)) +
  labs(title="2015-2016 NFL Offense/Defense Overview",x="Offensive YPG", y="Defensive YPG")  + scale_color_manual(values=color.scheme) +
  geom_text(aes(label=Team),size=3.0)  +
  theme_bw() + xlim(350, 460) + ylim(310,470) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.text=element_text(size=10), title=element_text(size=16)) +
  scale_size(range=c(5,17)) 
plt
```


Now let's remove the color legend and move the size legend to the bottom.

```{r fig.width=10, fig.height=10, echo=TRUE,message=FALSE, warning=FALSE} 
plt <- ggplot(data=nfl,aes(x=off_ypg,y=def_ypg)) + geom_point(aes(color=Team, size=ppg)) +
  labs(title="2015-2016 NFL Offense/Defense Overview",x="Offensive YPG", y="Defensive YPG")  + scale_color_manual(values=color.scheme) +
  geom_text(aes(label=Team),size=3.0)  + 
  theme_bw() + xlim(350, 460) +
  ylim(310,470) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.text=element_text(size=10), title=element_text(size=16)) +
  scale_size(range=c(5,17)) +
  guides(color=FALSE) +
  theme(legend.position = "bottom")
plt
```

Moving the legend freed up some space, so I'll slightly resize the labels. Lasly, let's deal with the cluttered overlapping points by adding some transparency, which is called `alpha` in most software. 

```{r fig.width=10, fig.height=10, echo=TRUE,message=FALSE, warning=FALSE}
plt <- ggplot(data=nfl,aes(x=off_ypg,y=def_ypg)) + geom_point(aes(color=Team, size=ppg),alpha=0.75) +
  labs(title="2015-2016 NFL Offense/Defense Overview",x="Offensive YPG", y="Defensive YPG")  + scale_color_manual(values=color.scheme) +
  geom_text(aes(label=Team),size=4.5)  + 
  theme_bw() + xlim(350, 460) + ylim(310,470) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())  + theme(axis.text=element_text(size=22), title=element_text(size=28)) + scale_size(range=c(8,23)) + guides(color=FALSE) + theme(legend.position = "bottom")
plt + coord_fixed(ratio = .5)
```

Looks pretty good!

Now that we did all that work, what can we learn from a graph like this? On such a plot, it's better to be low and right, meaning your defense doesn't give up much and your offense is producing well. Although many teams are sort of middle-of-the-road, the Broncos defense really stands out, and many would agree this was the driving force behind their Super Bowl win. Peyton Manning is one of the best to ever play the game, holding the all-time record for touchdown passes, passing yards, and overall wins. He is [surely destined for the Hall-of-Fame](http://broncoswire.usatoday.com/2016/03/06/peyton-manning-will-be-eligible-for-nfl-hall-of-fame-in-2021/); however, his performance in his final season was somewhat mediocre in comparison.  Luckily, the Broncos had Super Bowl MVP Von Miller who did [this](http://www.nfl.com/videos/nfl-game-highlights/0ap3000000634410/Von-Miller-highlights). Looks like defense made the difference this season.


